Fabric-Elasticity Relationships of Tibial Trabecular Bone are Similar in Osteogenesis Imperfecta and Healthy Individuals

### 2021-06-02 09_Article/Simon_2021_Bone_Draft_V01.tex
Finish tBMD analysis and prepare SBMS presentation

\begin{document}
\let\WriteBookmarks\relax
\def\floatpagepagefraction{1}
\def\textpagefraction{.001}
\shorttitle{Fabric-Elasticity in Osteogenesis Imperfecta}
\shortauthors{Simon et~al.}

\title[mode = title]{Fabric-Elasticity Relationships in Osteogenesis Imperfecta (OI)}

@@

\author[2,3]{Bettina Willie}

@@

\address[2]{Shriners Hospital for Children, Montreal, Canada}

\address[3]{McGill University, Montreal, Canada}

% Abbreviations
\abbreviations{OI, osteogenesis imperfecta;
			   HR-pQCT, high resolution peripheral quantitative tomography;
			   BMD, bone mass density;
			   ROI, region of interest;
			   BV/TV, bone volume over total volume;
			   SMI, structure model index;
			   MIL, mean intercept length;
			   FEA, finite element analysis
			   KUBC, kinematic uniform boundary condition;
			   PMUBC, periodicity-compatible mixed uniform boundary condition.}

@@

\begin{abstract}
	Osteogenesis Imperfecta (OI) is an inherited form of bone fragility. This disease, also called "brittle bone disease", is characterised by impaired synthesis of type I collagen, altered trabecular bone architecture and reduced bone mass that lead to fragile bones. High resolution peripheral computed tomography (HR-pQCT) is a powerful method to investigate bone morphology in the weight-bearing distal tibia and the 3D reconstructions can be exploited in either micro-finite element ($\mu$FE) or homogenised finite element (hFE) analysis for bone strength estimation. The hFE scheme uses local bone volume fraction (BV/TV) and anisotropy information (fabric) to compute healthy bone strength within a reasonable computation time using fabric-elasticity relationships. Thus, the aim of this study is to investigate fabric-elasticity relationships in OI trabecular bone compared to healthy controls.

	In this study, the morphology of distal tibias of 50 OI diagnosed people were compared to 120 healthy controls using second generation HR-pQCT. 
	Six regions of interest (ROIs) were selected per individual in a common anatomical region. A first age \& gender matching allowed to perform a morphometric analysis and compare the outcome with literature. Then, stiffness tensors of ROIs were computed by FEA and multiple linear regressions were performed on the Zysset-Curnier orthotropic model to compare the two groups using 5 parameters. After initial fits with all the samples of each group, the data were filtered according to a fixed threshold for a defined coefficient of variation (CV) assessing the ROI heterogeneity. Second fits were performed on these filtered data sets and then additional fits were done on BV/TV \& fabric anisotropy (DA) matched data to detect statistical differences between the two groups. This full and filtered data were in turn compared with previous results from $\mu$CT reconstructions obtained in other anatomical locations.

	In agreement with available literature, significantly lower BV/TV, trabecular number (Tb N), significantly higher trabecular spacing (Tb Sp), trabecular spacing standard deviation (Tb Sp SD), but no differences in trabecular thickness (Tb Th) were found between OI and controls. The stiffness of ROIs from OI bone reached lower values and the multilinear fabric-elasticity fits tended to overestimate the stiffness in the lower range. Filtering out highly heterogeneous ROIs removed these low stiffness ROIs and lead to similar correlation coefficients for both OI and healthy groups. Finally, the BV/TV \& DA matched data revealed no significant differences in fabric-elasticity parameters between OI and healthy individuals. Comparing with previous studies, the stiffness constants from 61 $\mu$m resolution HR-pQCT ROIs were lower than for 36 $\mu$m resolution $\mu$m CT images.

@@

Osteogenesis imperfecta (OI) is an inherited form of bone fragility \cite{Tournis2018}. It was estimated to concern about 1/13'500 births, less severe forms being not accounted in this estimation as they are recognized later in life \cite{Lindahl2015}. Therefore, OI is considered as a rare metabolic bone disorder. Sillence classification \cite{Sillence1979} allows to define four main types of OI based on the clinical phenotype:
\begin{itemize}
	\item Type I: less severe
	\item Type II: lethal at most shortly after birth
	\item Type III: most severe surviving form
	\item Type IV: intermediate severity
\end{itemize}
Also known as "brittle bone disease", OI is characterized by impaired synthesis of type I collagen, leading to such brittle and fragile bones \cite{LIM2017}.\\

Bone fragility in OI is complex and not totally understood despite the investigations at different hierarchical levels. Multiple studies show that DXA areal BMD (aBMD) tends to be lower in OI than compared to healthy individuals \cite{Folkestad2012,Lindahl2015,Scheres2018}. The microstructure is different as well. \citeauthor{Folkestad2012}\cite{Folkestad2012}, \citeauthor{Kocijan2015}\cite{Kocijan2015}, and \citeauthor{Rolvien2018}\cite{Rolvien2018} shown that BV/TV and trabecular number in OI is lower than for healthy controls. Trabecular spacing and inhomogeneity are higher for OI diagnosed people but the trabecular thickness is not significantly different. At the ECM level, a recent study shown that OI bone tends to present higher modulus, ultimate stress and post-yield behavior than healthy bone in compression \cite{Indermaur2021}.\\

High resolution peripheral quantitative tomography (HR-pQCT) scans allow to perform \textit{in vivo} assessment of trabecular architecture and volumetric BMD in distal radius and tibia \cite{Boutroy2005}. Moreover, the bony microstructure obtained from HR-pQCT can be used for finite element analysis (FEA) to predict mechanical properties \cite{Boutroy2008}. Homogenized FE (hFE) is a FEA scheme including volume fraction (BV/TV) and anisotropy information (fabric) of trabecular bone from HR-pQCT which to assess bone strength within reasonable computation time \cite{Pahr2009}. High correlations were found between patient-specific hFE and mechanical compression experiment of cadaveric samples \cite{Varga2011,AriasMoreno2019}. Thus, it could be legitimate to use hFE for OI patients bone strength estimation and fracture risk assessment. However, HR-pQCT-based FEA rely on fabric-elasticity relationships. Therefore, the present study aims to compare trabecular bone microstructure of healthy and OI diagnosed individuals and to investigate the hypothesis of similar fabric-elasticity relationships.

@@

\subsection{Subjects}
The healthy group include a total of 120 patients from a previous reproducibility study performed at the University Department of osteoporosis in Bern \cite{Schenk2020}. The sample is composed of 64 female and 56 male subjects aged between 20 and 92 years old with a median age of 26 [22 - 35] years. These subjects did not took any medication known to affect bone metabolism nor presented any prior osteoporosis fracture. The second group was scanned at the Shriners Hospital for Children and images were shared to our group by the McGill University in Montreal. This group is composed of 35 female 15 male individuals leading to 50 OI diagnosed subjects. The youngest and oldest patients are 19 and 69 year old, respectively. The median age is 44 [33 - 55] years. 35 subjects were diagnosed with type I OI, 2 with type III, and 13 with type IV.

\subsection{HR-pQCT}
HR-pQCT scans (XtremeCTII, SCANCO Medical, Brütisellen,
Switzerland) were performed at the distal tibia on all patients from both groups. People of the healthy group were scanned using an in-house protocol as described in \cite{Schenk2020} whereas OI subjects were scanned using the manufacturer's standard protocol. The main differences of the in-house protocol with the manufacturer's standard protocol are the following:
\begin{enumerate}
	\item The reference line is positioned at the proximal margin of the dense structure formed by the tibia plafond instead of  the subchondral endplate of the ankle joint (standard clinical section) \cite{Whittier2020}.
	\item Three stacks were scanned proximal to the reference line instead of one stack at 22.5 mm proximal to the reference line for standard clinical section \cite{Whittier2020}.
\end{enumerate}

These differences are shown in Figure \ref{01_ClinicalSections}. For both group, each stack consist of 168 voxels and a resolution of 61 \si{\micro}m in the three principal directions which lead to a thickness about 10.2 mm for each stack. Scanning settings were a voltage of 60 kVp, 900 μA, 100 ms integration time for the healthy group as well as for the OI group. For the healthy group, motion artefacts of first, middle and last slice were graded. The scale used start from 1 (no motion artefacts) to 5 (extreme motion artefacts) as proposed by the manufacturer \cite{Pialat2012}. The final grade of each scan was defined as the highest slice grade. For the OI group, as the scan consist of one stack, only one grade is attributed using the same scale as for the control group. Scans were then processed independently of their quality grading. A summary of the scans grading densities is shown in Figure \ref{01_MotionArtefacts}.

@@

\subsection{Image analysis}
The HR-pQCT scans were evaluated using the manufacturer's standard protocol. Then, mask segmented images were used for further analysis.\\

Six ROI were randomly selected in each scan. The conditions for ROI to be kept are no cortical bone inside and that it must contain trabecular bone. For the OI subjects, the stack was divided into two halves and the ROIs were selected to have the centers of three ROIs in the proximal half and three in the distal half. For the healthy people, the ROIs were selected in the more proximal stack uniquely, see Figure \ref{01_Healthy} stack n\textdegree 3. As for the OI subjects, the stack is halved and centers of three ROIs were selected in both half.\\

The ROI is defined as a cube of 5.3 mm side length. This size was chosen to correspond to the work of \citeauthor{Panyasantisuk2015}\cite{Panyasantisuk2015} and \citeauthor{Gross2013}\cite{Gross2013} which performed similar analysis with femur \si{\micro}CT scans. It was determined by \citeauthor{Zysset1998}\cite{Zysset1998} and \citeauthor{Daszkiewicz2017}\cite{Daszkiewicz2017} to be the optimal size to obtain accurate FEA results.\\

The morphological analysis of ROIs was performed using medtool (v4.5; Dr. Pahr Ingenieurs e.U., Pfaffstätten, Austria). The morphological parameters analyzed are: BV/TV, SMI, trabecular number (Tb. N.), trabecular thickness (Tb. Th.), trabecular spacing (Tb. Sp.), and the standard deviation of the trabecular spacing (Tb. Sp. SD). Moreover, ROI fabric was evaluated using MIL method \cite{Moreno2014}. The fabric tensor $\mathbf{M}$ is a positive-definite second-order tensor. It is build as shown in Equation \ref{Eq201} below:

@@

where $m_i$ are the eigenvalues of $\mathbf{M}$ and $\mathbf{M}_i$ are the dyadic product of the corresponding eigenvectors $\mathbf{m}_i$ \cite{Cowin1985,Harrigan1985}. The fabric is then independent of BV/TV and normalized with $tr(\mathbf{M}) = 3$. The fabric eigenvalues allow to compute the degree of anisotropy (DA) of the ROI by dividing the highest eigenvalue by the lowest one.\\

After ROI cleaning, i.e. deletion of unconnected region of bone material, an homogenized mechanical analysis was performed using \textsc{ABAQUS 6.14}. Each voxel of the cleaned ROI was converted to mesh using fully integrated linear brick elements (C3D8) with a stiffness $E$ of 10 000 MPa and a Poisson's ratio $\nu$ of 0.3. The simulation consisted of 6 independent load cases, 3 uni-axial and 3 shear cases, using KUBCs. KUBCs were used according to the work of \citeauthor{Panyasantisuk2015}\cite{Panyasantisuk2015}. Unlike PMUBCs, KUBCs do not require to rotate the ROI into fabric coordinate system which decrease the image quality. This homogenization process allows to calibrate the parameters of the Zysset-Curnier fabric-elasticity model \cite{Zysset1995}. This model builds the fourth order stiffness tensor $\mathbb{S}$ using the BV/TV or $\rho$, fabric information $\mathbf{M}$, three elasticity parameters $\lambda_0$, $\lambda_0$', and $\mu_0$, and two exponents, $k$ and $l$. The building of this tensor is shown in Equation \ref{Eq202}.\\

@@

Where $\otimes$ and $\overline{\underline{\otimes}}$ are the dyadic and symmetric product of second order tensors, respectively. The Zysset-Curnier model is built with the assumption of orthotropy and homogeneity. However, the trabecular structure is not perfectly homogeneous. In order to assess the ROI heterogeneity a so-called coefficient of variation (CV) is computed as presented in \cite{Panyasantisuk2015}: the ROI is divided into eight identical subcubes, BV/TV is computed for each subcube and the CV is defined as the ratio between the standard deviation of these BV/TV and the mean value, see Equation \ref{Eq203} below:

\begin{equation}
	CV = \frac{std(BV/TV_{subcubes})}{mean(BV/TV_{subcubes})}
	\label{Eq203}
\end{equation}

\subsection{Statistics}
The morphological parameters analyzed (BV/TV, Tb. N., Tb. Th., Tb. Sp., and Tb. Sp. SD, SMI, DA, and CV) were compared for both group. As the initial groups do not have similar distributions of age and sex, a matching was performed leading to identical mean and median age as well as identical gender distribution. Then, the selection of statistical test to perform was executed as follow:
\begin{enumerate}
	\item For each parameter, the median value between the six ROI of the same individual was computed. The median was preferred over the mean because it is less influenced by outliers.
	\item Normality of the distribution was assessed with QQ plot and Shapiro-Wilk test.
	\item If normality assumption was met, Bartlett test for equal variances was performed. Otherwise, data were log transformed to try to achieve normal distribution. If even after log transformation, normality assumption was not met, original data space was kept and Brown-Forsythe test was applied to assess the equal variance assumption.
	\item According to the previous results, t-test was performed if normal distribution and equal variance were met. If only the equal variances assumption was met, Mann-Whitney test was preferred. Finally, if none of these conditions could be assumed, a non-parametric permutation test was performed.
\end{enumerate}
The general significance level was set to 95\% for all tests. Confidence intervals in means difference was computed for t-tested variables to confirm p values. As Mann-Whitney test is performed on the median, only corresponding p value is given. Finally, non-parametric permutation test is less powerful but give an empirical 95\% exclusion range and a p value. If the difference in means belong to this exclusion range, it can be stated that group means are different with 95\% certainty.

\subsection{Fit to model}
The stiffness tensors obtained from the mechanical simulations were transformed into fabric coordinate system and projected onto orthotropy, leading to 12 components. The resulting orthotropic stiffness tensors were then used to perform a multiple linear regression on the Zysset-Curnier model. Standard linear models assume independent and identically distributed (iid) variables. As this assumption is violated by the fact that six ROIs are analyzed by individual, a linear mixed-effect model was preferred. This last model, shown in Equation \ref{Eq204} in Laird-Ware form \cite{Laird1982}, takes into account the non-independence of ROIs from the same individual. Detailed form of this model is presented in Appendix \ref{A1}.

@@

Where $S_{rc}$ is the $r$th row and $c$th column of the non-zero element of the orthotropic stiffness tensor $\mathbb{S}$ in Mandel notation \cite{MANDEL1965}, $X$ is a $12n$x$p$ design matrix containing the the BV/TV and fabric info of the $n$ ROIs and $\beta$ is a $p$x1 vector of fixed effects containing model parameters. $Z$ is a $12n$x$f$ design matrix which contains data with individual dependence and $\delta$ is a $f$x1 vector composed of random factors. Finally, $\epsilon$ is a $12n$x1 vector containing the regression residuals.\\

The linear regression was performed on both group (healthy and OI) separately. To improve the fit quality, the data sets were filtered. The aim here is to filter out ROIs whose are too far from the assumption of homogeneity. Therefore, analogously to the work of \citeauthor{Panyasantisuk2015}\cite{Panyasantisuk2015}, a fixed threshold for the CV was used. To simplify comparison, the same value of 0.263 was fixed as exclusion criterion. Besides, the relation between BV/TV and CV was assessed using Spearman's correlation coefficient. Furthermore, to compare the stiffness constants ($\lambda_0$, $\lambda_0'$, and $\mu_0$) between the groups, regression must be performed on identical value ranges. To do this, a matching was performed for BV/TV and DA to find corresponding control ROI for each OI in the filtered groups. Best correspondences were kept and duplicates were dropped. Finally, as the regression is performed in the log space, it is necessary to use identical exponent ($k$ and $l$) for both group, weighting identically BV/TV and DA between regressions, to compare stiffness constants. The exponents were determined by grouping healthy and OI for regression. Then a modified system is used to perform the fit on separated groups.\\

Another modification of the model is to add a regressor for the group variable (healthy or OI), i.e. add a column to the design matrix $X$ and a row to the parameter vector $\beta$. This modified model is compared to the original by analysis of covariance (ANCOVA) using the fixed-effects only to determine the statistical significance of the group. Implementation of this modification was performed according to \cite{Fox2016}. The detailed linear systems for each model discussed here are available in Appendix \ref{A1} and a summary of the data sets used for the different methods is shown in Table \ref{Table1}.\\

The regression was performed using the \textsc{statsmodels} package from \textsc{Python 3.6}. Regression quality was assessed using the adjusted Pearson correlation coefficient squared ($R^2_{adj}$) and relative error between the orthotropic observed and the predicted tensor using norm of fourth-order tensors ($NE$), see Equation \ref{Eq205} and \ref{Eq206}.

>>>>

\begin{document}
\let\WriteBookmarks\relax
\def\floatpagepagefraction{1}
\def\textpagefraction{.001}
\shorttitle{Fabric-Elasticity Relationships in Osteogenesis Imperfecta}
\shortauthors{Simon et~al.}

\title[mode = title]{Fabric-Elasticity Relationships of Tibial Trabecular Bone are Similar in Osteogenesis Imperfecta and Healthy Individuals}

@@

\author[2,3]{Bettina M. Willie}

@@

\address[2]{Research Centre, Shriners Hospital for Children, Montreal, Canada}

\address[3]{Departement of Pediatric Surgery, McGill University, Montreal, Canada}

% Abbreviations
\abbreviations{OI, osteogenesis imperfecta;
			   HR-pQCT, high resolution peripheral quantitative computed tomography;
			   BMD, bone mass density;
			   ROI, region of interest;
			   BV/TV, bone volume over total volume;
			   SMI, structure model index;
			   MIL, mean intercept length;
			   FEA, finite element analysis
			   KUBC, kinematic uniform boundary condition;
			   PMUBC, periodicity-compatible mixed uniform boundary condition.}

@@

\begin{abstract}
	Osteogenesis Imperfecta (OI) is an inherited form of bone fragility, also called "brittle bone disease". It is characterised by impaired synthesis of type I collagen, altered trabecular bone architecture and reduced bone mass. High resolution peripheral computed tomography (HR-pQCT) is a powerful method to investigate bone morphology of extremities including the weight-bearing distal tibia. The resulting 3D reconstructions can be used as a basis of micro-finite element (\si{\micro}FE) or homogenised finite element (hFE) models for bone strength estimation. The hFE scheme uses homogenized local bone volume fraction (BV/TV) and anisotropy information (fabric) to compute healthy bone strength within a reasonable computation time using fabric-elasticity relationships. Thus, the aim of this study is to investigate fabric-elasticity relationships in OI trabecular bone compared to healthy controls.

	In this study, the morphology of distal tibiae from 50 adults with OI were compared to 120 healthy controls using second generation HR-pQCT. Six cubic regions of interest (ROIs) were selected per individual in a common anatomical region. A first matching was performed by selecting similar individuals to obtain identical mean and median age and gender distribution in the OI and healthy control group. It allowed to perform a morphometric analysis and compare the outcome with literature. Then, stiffness tensors of ROIs were computed using hFEA and multiple linear regressions were performed on the Zysset-Curnier orthotropic model. The regressions allowed to compare the two groups using 5 parameters. An initial fit was performed on both the OI group and the healthy control group using all the ROIs extracted. Then, data were filtered according to a fixed threshold for a defined coefficient of variation (CV) assessing the ROI heterogeneity and second fits were performed on these filtered data sets. These full and filtered data were in turn compared with previous results from \si{\micro}CT reconstructions obtained in other anatomical locations. Finally, the ROIs of both group were matched according to their BV/TV and fabric anisotropy (DA). Fits were performed again using these matched data to detect statistical differences between the two groups.

	Compared to healthy controls, we found the OI samples to have significantly lower BV/TV and trabecular number (Tb.N.), significantly higher trabecular separation (Tb.Sp.) and trabecular spacing standard deviation (Tb.Sp.SD), but no differences in trabecular thickness (Tb.Th.). These results are in agreement to literature. The stiffness of ROIs from OI bone reached lower values compared to healthy controls and the multilinear fabric-elasticity fits tended to overestimate the stiffness in the lower range. The filtering of highly heterogeneous ROIs removed these low stiffness ROIs and lead to similar correlation coefficients for both OI and healthy groups. Finally, the BV/TV and DA matched data revealed no significant differences in fabric-elasticity parameters between OI and healthy individuals. Compared to previous studies, the stiffness constants from the 61 \si{\micro}m resolution HR-pQCT ROIs were lower than for the 36 \si{\micro}m resolution \si{\micro}CT ROIs.

@@

Osteogenesis imperfecta (OI), also commonly known as "brittle bone disease", is an inherited form of bone fragility \cite{Tournis2018}. OI prevalence is estimated at about  1/13,500 of births, less severe forms being not accounted in this estimation as they are recognized later in life \cite{Lindahl2015}. Therefore, OI is considered as a rare metabolic bone disorder. In most cases, OI is caused by mutations in genes encoding type I collagen (COL1A1 and COL1A2), leading to brittle and fragile bones \cite{LIM2017}, as well as deformed geometry and size in some cases. OI can be categorized according to disease severity \cite{Mortier2019} into:
\begin{itemize}
	\item Type I: mild
	\item Type II: perinatally lethal
	\item Type III: most severe surviving form
	\item Type IV: intermediate severity
\end{itemize}

Bone fragility in OI is complex and not totally understood, despite the investigations at different hierarchical levels. Multiple studies show that DXA areal bone mineral density (aBMD) tends to be lower in OI compared to healthy individuals \cite{Folkestad2012,Lindahl2015,Scheres2018}.  \citeauthor{Folkestad2012}\cite{Folkestad2012}, \citeauthor{Kocijan2015}\cite{Kocijan2015}, and \citeauthor{Rolvien2018}\cite{Rolvien2018} have shown that the microstructure is different as well, namely bone volume fraction (BV/TV) and trabecular number (Tb.N.) in OI bone is lower than for healthy controls. Trabecular separation (Tb.Sp.) and inhomogeneity (Tb.Sp.SD) are higher for individuals with OI but the trabecular thickness (Tb.Th.) is not significantly different. At the ECM level, a recent study showed that, in compression, OI bone tends to present higher modulus, ultimate stress and post-yield behavior than healthy bone, mostly affected by the higher degree of mineralization of OI bone \cite{Indermaur2021}.\\

High resolution peripheral quantitative computed tomography (HR-pQCT) scans allow one to perform \textit{in vivo} assessment of cortical and trabecular architecture and volumetric bone mineral density (BMD) in the distal radius and distal tibia \cite{Boutroy2005}. Moreover, the image obtained from HR-pQCT can be used for finite element analysis (FEA) to predict mechanical properties \cite{Boutroy2008}. Homogenized finite element (hFE) is based on BV/TV and anisotropy information (fabric) from the HR-pQCT scan that can be used to assess bone strength within a reasonable computation time \cite{Pahr2009}. High correlations were found between patient-specific hFE and mechanical compression experiments of freshly frozen human samples at the distal radius \cite{Varga2011,AriasMoreno2019}. Thus, it could be legitimate to use hFE for OI patient's bone strength estimation and potentially fracture risk assessment. However, HR-pQCT-based FEA relies on fabric-elasticity relationships. Therefore, the present study aims to compare trabecular bone microstructure of healthy and OI bone samples and to investigate the hypothesis of similar fabric-elasticity relationships.

@@

\subsection{Participants}
The healthy group included a total of 120 patients from a previous reproducibility study performed at the University Department of osteoporosis in Bern \cite{Schenk2020}. The group was composed of 64 females and 56 males aged between 20 and 92 years old with a mean age of 32 $\pm$ 15 years. These subjects had not taken any medication known to affect bone metabolism nor presented with any prior osteoporosis fracture. The second group was scanned as part of the ASTEROID study at different locations in Canada, namely at the Shriners Hospital for Children-Canada. The study coordination was done by the McGill University in Montreal. This group was composed of 35 females and 15 males with confirmed diagnosis of OI Type I, III or IV. There were 35 patients diagnosed with OI type I, 2 with type III, and 13 with type IV. The participants of the OI group were aged between 19 and 69 years old with a mean age of 44 $\pm$ 14 years.

\subsection{HR-pQCT}
HR-pQCT scans (XtremeCTII, SCANCO Medical, Brütisellen,
Switzerland) were performed at the distal tibia on all patients from both groups. Participants in the healthy group were scanned using an in-house protocol as described in \cite{Schenk2020}. Namely, the reference line was positioned at the proximal margin of the dense structure formed by the tibia plafond and three stacks were scanned proximal to this reference line, see Figure \ref{01_Healthy}. On the other hand, participants in the OI group were scanned using the manufacturer's standard protocol i.e. the reference line is placed at the subchondral endplate of the ankle joint and one stack was scanned at 22.5 mm proximal to the reference line (standard clinical section) \cite{Whittier2020}, see Figure \ref{01_OI}. The scanned region according to the two different protocol are shown side by side in Figure \ref{01_ClinicalSections}. For both the healthy and the OI groups, each stack consisted of 168 slices and a voxel resolution of 61 \si{\micro}m in the three principal directions. This led to a thickness of roughly 10.2 mm for each stack. Standardized scanning settings were used (voltage of 60 kVp, 900 μA, 100 ms integration time) for the healthy group as well as for the OI group. For the healthy group, motion artefacts of first, middle and last slice (i.e slices number 1, 252, and 504) were graded on a scale of 1 (no motion artefacts) to 5 (extreme motion artefacts), as proposed by \citeauthor{Pialat2012} \cite{Pialat2012}. The final grade of each scan was defined as the highest slice grade. For the OI group, as the scan consisted of one stack, only one grade was attributed using the same scale as for the control group. Scans were then processed independently of their quality grading. A summary of the scans grading is shown in Figure \ref{01_MotionArtefacts}.

@@

\subsection{Image analysis}
The HR-pQCT scans were evaluated using the manufacturer's standard protocol. Briefly, an automatic contouring algorithm was applied to define the periosteal contour of the tibia (masking) and a threshold was applied for segmentation of cortical bone (450 mgHA/cm\textsuperscript{3}) and trabecular bone (320 mgHA/cm\textsuperscript{3}). Then, mask segmented images were used for further analysis.\\

In general, six cubic ROIs were selected at random position in each scan in a defined area. Each ROI had to contain trabecular bone, but no cortical bone. For the healthy group, the ROIs were selected in the more proximal stack uniquely (see Figure \ref{01_Healthy} stack number 3) to be at the same anatomical location as for the OI group. Then for both the healthy group and the OI group, the stack was divided into two halves and the ROIs were selected to have the centres of three ROIs in the proximal half and three in the distal half. For one individual diagnosed with OI type III it was not possible to extract any ROI as there was no enough trabecular bone. This led to 720 healthy ROIs from 120 individuals and 294 OI ROIs from 49 individuals.\\

The ROI was defined as a cube of 5.3 mm side length. This size was in agreement with the work of \citeauthor{Panyasantisuk2015}\cite{Panyasantisuk2015} and \citeauthor{Gross2013}\cite{Gross2013}, who performed similar analysis with femur \si{\micro}CT scans. It was determined by \citeauthor{Zysset1998}\cite{Zysset1998} and \citeauthor{Daszkiewicz2017}\cite{Daszkiewicz2017} to allow one to have a relative homogeneity of trabecular tissue within the ROI leading to accurate \si{\micro}FE results with a minimal computational cost.\\

After ROI cleaning, i.e. deletion of unconnected region of bone material, the morphological analysis of ROIs was performed using medtool (v4.5; Dr. Pahr Ingenieurs e.U., Pfaffstätten, Austria). The morphological parameters analyzed were: BV/TV, structural model index (SMI), trabecular number (Tb.N.), trabecular thickness (Tb.Th.), trabecular separation (Tb.Sp.), and the standard deviation of the trabecular spacing (Tb.Sp.SD). Moreover, ROI tissue bone mineral density (tBMD) and fabric was evaluated. The fabric tensor $\mathbf{M}$ was computed using mean intercept length (MIL) method \cite{Moreno2014}. It is a positive-definite second-order tensor built as shown in Equation \ref{Eq201} below:

@@

where $m_i$ are the eigenvalues of $\mathbf{M}$ and $\mathbf{M}_i$ are the dyadic product of the corresponding eigenvectors $\mathbf{m}_i$ \cite{Cowin1985,Harrigan1985}. The fabric tensor is independent of BV/TV and normalized with $tr(\mathbf{M}) = 3$. The fabric eigenvalues allow to compute the degree of anisotropy (DA) of the ROI by dividing the highest eigenvalue by the lowest one. Figure \ref{01_FabricExample} shows an example of a typical ROI with the visualization of its fabric tensor.\\

\begin{figure}[h!]
	\centering
	\includegraphics[width=\linewidth, trim= 0 0 0 100]
	{Pictures/01_FabricExample}
	\caption{Typical ROI with the visualization of its fabric tensor using MIL method. Eigenvectors of the fabric tensor define its orientation and eigenvalues set lengths of the ellipsoid radii. DA is the ratio between the highest and the lowest eigenvalue.}
	\label{01_FabricExample}
\end{figure}

A \si{\micro}FE mechanical analysis was performed using \textsc{ABAQUS 6.14}. In brief, each voxel of the cleaned ROI was converted to a fully integrated linear brick elements (C3D8) using a direct voxel conversion approach. Then, a stiffness $E$ of 10,000 MPa and a Poisson's ratio $\nu$ of 0.3 were assigned. The homogenization process consisted of 6 independent simulations of different load cases, 3 uni-axial and 3 simple shear cases, using kinematic uniform boundary conditions (KUBCs) \cite{Panyasantisuk2015}. Unlike periodicity-compatible mixed uniform boundary conditions (PMUBCs), KUBCs do not require one to rotate the ROI into fabric coordinate system. Such rotation would potentially decrease image quality. The homogenization process allowed to calculate the components of the stiffness tensor and to calibrate the parameters of the Zysset-Curnier fabric-elasticity model \cite{Zysset1995}. This model builds the fourth order stiffness tensor $\mathbb{S}$ using the BV/TV or $\rho$, fabric tensor $\mathbf{M}$, three elasticity parameters $\lambda_0$, $\lambda_0$', and $\mu_0$, and two exponents, $k$ and $l$, as shown in Equation \ref{Eq202}.\\

@@

Where $\otimes$ and $\overline{\underline{\otimes}}$ are the dyadic and symmetric product of second order tensors, respectively. To express the stiffness tensor obtained from the homogenization process with the Zysset-Curnier model, it had to be transformed into the fabric coordinate system using a coordinates transformation formula (see Equation \ref{Eq203p}) and projected onto orthotropy, leading to 12 components.

\begin{equation}
	\mathbb{S}_{ijkl}' = Q_{im}Q_{jn}Q_{ko}Q_{lp} \mathbb{S}_{mnop}
	\label{Eq203p}
\end{equation}

Where $\mathbb{S}'$ and $\mathbb{S}$ are the transformed and the original stiffness tensor respectively and $Q$ is the orthogonal matrix that maps the original coordinate system into the new one (fabric). The Zysset-Curnier model is built with the assumption of orthotropy and homogeneity. However, the trabecular structure is not perfectly homogeneous. In order to assess the ROI heterogeneity, a coefficient of variation (CV) is computed according to \citeauthor{Panyasantisuk2015}\cite{Panyasantisuk2015}: the ROI is divided into eight identical sub-cubes and BV/TV is computed for each of them. The CV is defined as the ratio between the standard deviation of these BV/TV and the mean value (Equation \ref{Eq203}).

\begin{equation}
	CV = \frac{SD(BV/TV_{subcubes})}{mean(BV/TV_{subcubes})}
	\label{Eq203}
\end{equation}

\subsection{Statistics}
The analyzed morphological parameters (BV/TV, Tb.N., Tb.Th., Tb.Sp., and Tb.Sp.SD, SMI, DA, and CV) were compared between the healthy and the OI groups. As the groups do not have similar distributions of age and sex, a matching was performed by selecting similar individuals leading to identical mean and median age as well as identical gender distribution. For each parameter, the median value between the six ROIs from the same individual was computed. The median was preferred over the mean because it is less influenced by outliers. Normality of the distribution was assessed with QQ plot and Shapiro-Wilk test. CV had to be log-transformed to meet normal distribution assumption. Then, equal variances was assessed using Bartlett test or Brown-Forsythe test according to the normality distribution of the data. According to the normality and equal variances assumptions, t-test, Mann-Whitney test or a non-parametric permutation test was performed. The general significance level was set to 95\% for all tests. Confidence intervals was computed for t-tested variables to confirm a difference in both groups means. As Mann-Whitney tests are performed on the median, only the corresponding p-value is presented. Finally, non-parametric permutation tests are less powerful but give an empirical 95\% exclusion range and a p-value. If the difference in means belong to this exclusion range, it can be stated that group means are different with 95\% certainty.

\subsection{Linear Regression}
The orthotropic stiffness tensors obtained after transformation onto fabric coordinate system were then used to perform a multiple linear regression with the Zysset-Curnier model. Standard linear models assume independent and identically distributed (iid) variables. As this assumption was violated by the fact that six ROIs were analyzed per individual, a linear mixed-effect model was preferred. This last model, shown in Equation \ref{Eq204} in Laird-Ware form \cite{Laird1982}, considered the non-independence of ROIs from the same individual. A more detailed form of this model is presented in Appendix \ref{A1}.

@@

Where $S_{rc}$ is the $r$th row and $c$th column of the non-zero element of the orthotropic stiffness tensor $\mathbb{S}$ in Mandel notation \cite{MANDEL1965}, $X$ is a $12n$x$p$ design matrix containing the the BV/TV and fabric info of the $n$ ROIs and $\beta$ is a $p$x1 vector of fixed effects containing model parameters. $Z$ is a $12n$x$f$ design matrix which contains data with individual dependence and $\delta$ is a $f$x1 vector composed of random factors. Finally, $\epsilon$ is a $12n$x1 vector containing the regression residuals. As $k$ and $l$ are exponents, the regression was performed on the log space.\\

The linear regression was performed on both group (heal\-thy and OI) separately. To improve the fit quality, the data sets were filtered. The aim here was to filter out ROIs violating the assumption of homogeneity. Therefore, analogously to the work of \citeauthor{Panyasantisuk2015}\cite{Panyasantisuk2015}, a fixed threshold for the CV was used. To simplify comparison, we used the same value 0.263 as exclusion criterion \cite{Panyasantisuk2015}. Then, the relation between BV/TV and CV was assessed using Spearman's correlation coefficient. To compare the stiffness constants ($\lambda_0$, $\lambda_0'$, and $\mu_0$) between the groups, regression must be performed on identical value ranges. Therefore, a matching was performed for BV/TV and DA to find corresponding control ROIs for each OI in the filtered groups. Best correspondences were kept and duplicates were dropped. Finally, as the regression was performed in the log space, slight differences in the exponents ($k$ and $l$) would lead to important variation of the stiffness constants ($\lambda_0$, $\lambda_0'$, and $\mu_0$) so it was necessary to use identical exponents  for both groups, weighting identically BV/TV and DA between regressions. The exponents were determined by grouping healthy and OI for regression. Then a modified system was used to perform the fit on separated groups, see Appendix \ref{A1}, Equation \ref{EqA11}.\\

Another modification of the model was to add a regressor for the group variable (healthy or OI), i.e. add a column to the design matrix $X$ and a row to the parameter vector $\beta$. This modified model is compared to the original by analysis of covariance (ANCOVA) using the fixed-effects only to determine the statistical significance of the group. Implementation of this modification was performed according to \cite{Fox2016}. A similar mixed-effect model was used to analyze the relation between tBMD and BV/TV and the significance of the group, see Appendix \ref{A1} Equation \ref{EqA14}. The model used the BV/TV and the group (healthy or OI) as fixed variables and the individual as random variable. Moreover, to test the hypothesis of no interaction between the BV/TV and the group, i.e. the group has no significant influence on the tBMD versus BV/TV slope, the model was modified to add the interaction regressor (BV/TV x Group). The detailed linear systems for each model discussed here are available in Appendix \ref{A1} and a summary of the data sets used for the different methods is shown in Table \ref{Table1}.\\
